---
title: "US crime"
output: html_document
date: "2023-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

```{r message=F}
df <- USArrests
library(dplyr)
library(stats)
library(factoextra)
library(cluster)
library(usmap)
library(tidyverse)
df<-na.omit(df)
```

## Distance
```{r}
set.seed(123)
df.scale <- scale(df)
diss.matrix <- dist(df.scale, method = "euclidean")
```

## Clustering
```{r}
set.seed(123)
clusters.h<- hclust(diss.matrix, method = "complete" )
plot(clusters.h, cex = 0.6, hang = -1)
```

## Finding optimal number of clusters

```{r}
hclusCut <- function(x, k) list(cluster = cutree(hclust(dist(x, method="euclidian"), method="complete"), k=k))
gap_stat <- clusGap(df.scale, FUN = hclusCut, K.max = 20, B = 100)
fviz_gap_stat(gap_stat)
```


This means we choose 4 clusters

## Investigating the clusters
```{r}
clusters.h.4<-cutree(clusters.h, k=4)
table(clusters.h.4)
```
We can see that there is not an equal split of states in each cluster

We add the cluster to the original data:

```{r}
df <- df %>% mutate(clusters = as.character(clusters.h.4)) %>% 
    mutate(state = names(clusters.h.4))

```

We can visualize the dendrogram with the cut of 4 clsuters
```{r, warning=F}
fviz_dend(clusters.h, rect=TRUE,k = 4)
```

Creating the clusters diagram which visualizes the clusters with PCA
```{r}
set.seed(123)
clusters.hcut<-hcut(df.scale, k=4, hc_func="hclust", hc_method="complete", hc_metric="euclidian")
fviz_cluster(clusters.hcut)
```

Investigating the differences of the features for each cluster:
```{r}
df %>% pivot_longer(-c(clusters,state)) %>% 
    ggplot(aes(y = value, x = clusters,
               color = clusters))+
    geom_boxplot()+
    facet_grid(~name,scales = "free_y") +
    scale_y_continuous(trans = "log10")+
    theme_bw(15)
```


Visualizing it geographically
```{r}
df.red <- df %>% dplyr::select(clusters,state)
colnames(df.red) <- c("clusters","full")
map_df <- us_map() %>% full_join(df.red, by = "full")
map_df <- na.omit(map_df)
ggplot(map_df, aes(x = x, y = y,
                   color = clusters))+
    geom_point(size = .8)+
    theme_bw()

```